Normalization by Evaluation for Sized Dependent Types

### 2017-02-20 nbe/icfp17.tex
[ ICFP 17] Second fund. thm.

\begin{lemma}[Resurrection]
\label{lem:res} \bla
\begin{enumerate}
\item If\/ $\der \Gamma$ then $\der \resurrect\Gamma$ and $\resurrect \Gamma \der \sid : \Gamma$.
  % Consequently, if\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
  % Further, if $\Gamma \der J : \resurrect\Delta$ then $\Gamma \der J : \Delta$.
\item If\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
\item If\/ $\Gamma \der J : \resurrect \Delta$ and $\der \Delta$ then $\Gamma \der J : \Delta$.
\item If\/ $\Gamma \der \sigma : \Delta$ then $\resurrect\Gamma \der \sigma : \resurrect\Delta$.
\end{enumerate}
\end{lemma}

@@

@@

Validity judgements.
\[
\begin{array}{l@{\quad}l@{\quad}l}
  \valid () & \defas & \mtrue \\
  \valid \eext \Gamma \Size & \defas & \valid \Gamma \\
  \valid \cext \Gamma T     & \defas & \valid \Gamma \mand \resurrect\Gamma \valid T
\\[2ex]
  \Gamma \valid s & \defas & \valid \Gamma \\
%  \Gamma \valid s & \defas & \mtrue \\
  \Gamma \valid T & \defas & \Gamma \valid T = T \\
  \Gamma \valid T = T' & \defas & \Gamma \valid T = T' : s \mforsome s %\Set[\ell] \mforsome \ell
\\[2ex]
  \Gamma \valid t : T & \defas & \Gamma \valid t = t : T \\
  \Gamma \valid t = t' : T & \defas & %\valid \Gamma \mand
    % \resurrect\Gamma \valid T \mand
    \valid \cext \Gamma T \mand
    \LR {\Gamma'} {t \sigma} {T \tau} {t' \eta} {T \rho}
    \mforall \LRSub {\Gamma'} \sigma \tau \Gamma \eta \rho
\\[2ex]
  \Gamma \valid \sigma : \Delta & \defas & \Gamma \valid \sigma = \sigma \eeq \sigma : \Delta \\
  \Gamma \valid \sigma_1 = \sigma_2 \eeq \tau : \Delta & \defas &
    \valid \Gamma \mand \valid \Delta \mand
    \LRSub {\Gamma'} {\sigma_1 \sigma'} {\tau \tau'} \Delta {\sigma_2 \eta'} {\tau \rho'}
    \mforall \LRSub {\Gamma'} {\sigma'} {\tau'} \Gamma {\eta'} {\rho'}
\end{array}
\]

\begin{theorem}[Fundamental theorem]
  \bla
  \begin{enumerate}
  \item
  If $\der \Gamma$ then $\valid \Gamma$.
  \item
  If $\Gamma \der J$ then % $\valid \Gamma$ and
  $\Gamma \valid J$
  \end{enumerate}
\end{theorem}
\begin{proof}
Simultaneously, by induction on the derivation.

\begin{caselist}
\nextcase $\forall$-introduction.
\[
  \ru{\erext \Gamma \Size \der t = t' : T
    }{\Gamma \der \lambda t = \lambda t' : \forall T}
\]
First $\valid \Gamma$ follows from the induction hypothesis $\valid {\erext \Gamma \Size}$.
To show $\resurrect\Gamma \valid \forall T$,
assume $\eta = \eta' \eeq \rho \in \resurrect \Gamma$
and show $(\forall T)\eta = (\forall T)\eta' \in \SET[\ell]$ for some $\ell$.
To this end, assume $\alpha \in \SIZE$ and show $T(\eta,\alpha) = T(\eta',\alpha) \in \SET[\ell]$.
Note that $(\eta,\alpha) = (\eta',\alpha) \eeq (\rho,\alpha) \in
\resurrect {\erext \Gamma \Size} = \cext {\resurrect\Gamma}\Size$,
thus, we can instantiate the induction hypothesis and obtain our goal.

\nextcase $\forall$-elimination.
\[
  \ru{\Gamma \der t = t' : \forall T \qquad
      \Gamma \der a, a' : \Size \qquad
      \resurrect\Gamma \der b : \Size
    }{\Gamma \der t \ann a = t' \ann {a'} : T[b]}
\]
First, $\valid \Gamma$ follows by induction hypothesis.
For goal $\resurrect \Gamma \valid T[b]$,
assume $\eta = \eta' \eeq \rho \in \resurrect\Gamma$ and show
$T[b]\eta = T[b]\eta' \in \SET[\ell]$ for some $\ell$.
By induction hypothesis,  $b\eta = b\eta' \in \SIZE$.
% , thus,
% $(\eta,b\eta) = (\eta',b\eta') \eeq (\rho,b\eta) \in \cext{\resurrect\Gamma}\Size$
By another induction hypothesis, $(\forall T)\eta = (\forall T)\eta' \in \SET[\ell]$,
which by definition entails our goal $T(\eta, b\eta) = T(\eta', b\eta') \in \SET[\ell]$.

>>>>

\begin{lemma}[Resurrection]
\label{lem:res} \bla
\begin{enumerate}
\item $\der \Gamma$ iff $\der \resurrect\Gamma$;  then $\resurrect \Gamma \der \sid : \Gamma$.
  % Consequently, if\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
  % Further, if $\Gamma \der J : \resurrect\Delta$ then $\Gamma \der J : \Delta$.
\item If\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
\item If\/ $\Gamma \der J : \resurrect \Delta$ % and $\der \Delta$
  then $\Gamma \der J : \Delta$.
\item If\/ $\Gamma \der \sigma : \Delta$ then $\Gamma \der \sigma : \resurrect\Delta$.
% then $\resurrect\Gamma \der \sigma : \resurrect\Delta$.
\end{enumerate}
\end{lemma}

@@

\begin{lemma}[Properties of the logical relation for substitution]
  \label{lem:wklrsub}
  Let $\LRSub \Gamma \sigma \tau \Delta \eta \rho$.
  \begin{enumerate}
  \item Weakening:
  If $\sigma' : \Gamma' \leq \Gamma$ then
  $\LRSub{\Gamma'}{\sigma\sigma'}{\tau\sigma'}{\Delta}\eta\rho$.
  \item Resurrection:
  Let $\LRSub {\Gamma} \tau \tau {\resurrect\Delta} \rho \rho$.
  \end{enumerate}
\end{lemma}

@@

Validity judgements.
\[
\begin{array}{l@{\quad}l@{\quad}l}
  \Gamma \valid t : T & \defas & %\valid \Gamma \mand
    % \resurrect\Gamma \valid T \mand
    \valid \cext \Gamma T \mand
    \LR {\Gamma'} {t \sigma} {T \tau} {t \eta} {T \rho}
    \mforall \LRSub {\Gamma'} \sigma \tau \Gamma \eta \rho
\\[1ex]
  \Gamma \valid \sigma : \Delta & \defas &
    \LRSub {\Gamma'} {\sigma \sigma'} {\sigma \tau'} \Delta {\sigma \eta'} {\sigma \rho'}
    \mforall \LRSub {\Gamma'} {\sigma'} {\tau'} \Gamma {\eta'} {\rho'}
\end{array}
\]

\begin{theorem}[Fundamental theorem of typing]
  \bla
  \begin{enumerate}
  \item
  If $\Gamma \der t : T$ then $\Gamma \valid t : T$.
  \item
  If $\Gamma \der \sigma : \Delta$ then
  $\Gamma \valid \sigma : \Delta$
  \end{enumerate}
\end{theorem}
\begin{proof}
Each by induction on the derivation.

\nextcase $\forall$-introduction.
\[
  \ru{\erext \Gamma \Size \der t : T
    }{\Gamma \der \lambda t : \forall T}
\]
Assume $\LRSub {\Delta} {\sigma} {\tau} \Gamma {\eta} {\rho}$ and show
$\LR \Delta {(\lambda t)\sigma} {(\forall T)\tau} {(\lambda t)\eta} {(\forall T)\rho}$.
To this end, assume a weakening $\sigma' : \Delta' \leq \Delta$ and
size expressions $\Delta' \der a : \Size$ and
$\resurrect{\Delta'} \der b : \Size$ and size values $\alpha,\beta$ with
$\LRSize {\Delta'} b \beta$ and show
$\LR {\Delta'} {t(\sigma\sigma',a)} {T(\tau\sigma',b)} {t(\eta,\alpha)} {T(\rho,\beta)}$.
This follows from the induction hypothesis, since
$\LRSub {\Delta'} {(\sigma\sigma',a)} {(\tau\sigma',b)} {\erext \Gamma \Size} {(\eta,\alpha)} {(\rho,\beta)}$
by Lemma~\ref{lem:wklrsub} and substitution extension.

\nextcase $\forall$-elimination.
\[
  \ru{\Gamma \der t : \forall T \qquad \Gamma \der a : \Size \qquad \resurrect \Gamma \der b : \Size
    }{\Gamma \der t \ann a : T[b]}
\]
Assume $\LRSub {\Delta} {\sigma} {\tau} \Gamma {\eta} {\rho}$ and show
$\LR \Delta {t \ann a \sigma} {T[b]\tau} {t \ann a \eta} {T[b]\rho}$.
By induction hypothesis we have $\LR \Delta {t \sigma} {(\forall T)\tau} {t \eta} {(\forall T) \rho}$.
It suffices to show
\begin{enumerate}
\item $\Delta \der a\sigma : \Size$
%\item $\resurrect\Delta \der b\tau : \Size$.
 and $a\eta \in \SIZE$: follows by induction hypothesis on $\Gamma \der a : \Size$.
%\item $b\rho \in \SIZE$.
\item $\LRSize {\resurrect \Delta} {b\tau} {b\rho}$: follows from induction hypothesis on $\resurrect \Gamma \der b : \Size$ via $\LRSub {\resurrect \Delta} \tau \tau {\resurrect\Gamma} \rho \rho$ (from Lemma~\ref{lem:wklrsub}).
\end{enumerate}
